\(Q\)-Boson model and relations with integrable hierarchies

### 2024-04-29 draft.tex
finish paper

\begin{abstract}
In this work, we study aspects of the \(Q\)-boson model, and 
its relations with integrable hierarchies. 
\end{abstract}

@@

Additionally, we explore other correlation functions within this model
and examine how they fit into the classical integrable context. It is
noteworthy that even when the object is not a tau function itself, it
may have an interesting expansion in terms of different classes of
Schur polynomials, highlighting some of its combinatorial
properties. We also investigate some of these expansions in this
paper.

@@

Furthermore, according to Zabrodin~\cite{Zabrodin:2010ii}, see
citations therein, we also know that under the rescaling \(t_k \to
T_k/ \hbar\), \(t'_k \to T_{-k}/ \hbar\) and \(N = T_0/ \hbar\), we
obtain the dispersion limit tau function
\begin{equation}
  F_0(\bm{T}) = \log S_{T_0}(\bm{T}, \bm{T}') + \mathcal{O}(\hbar)\; ,
\end{equation}
that is a free energy, from the viewpoit of the matrix integral
partition function.

@@

% Ablowitz-Ladik and Toda hierarchy - reduction -- understands this problem

%Let us now consider the fermionic construction of this tau function. We write 
%the limit \(N,M\to \infty\) of the scalar product
%\begin{subequations}
%\begin{equation}
%  S_Q(\bm{T}, \bm{t}') = \sum_{\mu} s_\mu(\bm{T}) s_\mu(\bm{t}')
%\end{equation}
%that is quite quite trivial from the viewpoint of the free fermion construction, 
%\begin{equation}
%  S_Q(\bm{T}, \bm{t}') = \bra{\bm{0}} e^{\bm{J}_+(\bm{T})} e^{\bm{J}_-(\bm{t}')} \ket{\bm{0}}\; .
%\end{equation}
%
%Moreover, we also know that
%\begin{equation}
%  S_Q(\bm{t}, \bm{t}') = \sum_{\mu\nu} c_{\mu\nu} s_\mu(\bm{t}) s_\mu(\bm{t}')
%\end{equation}
%where \(n t = \sum_j x^n_j\) and \(n t' = \sum_j y^n_j\) are Miwa coordinates, then  
%\begin{equation}
%  S_Q(\bm{t}, \bm{t}') = \bra{\bm{0}} e^{J_+(\bm{t})} G_Q  e^{J_-(\bm{t}')} \ket{\bm{0}}
%\end{equation}
%and \(G_Q \in GL(\infty)\) that depends on \(Q\). We could also think of \(G_Q\)
%as a twist
%\(G_Q = e^{Q H} G e^{-Q H}\), with the action of \(H\in \mathfrak{gl}(\infty)\)
%on ordinary fermions generating the Q-fermions. 
%\marginpar[left]{\tiny Investigate this\\ twists}
%\end{subequations}
%
%It seems that we have some different representations for this tau function, 
%and it is worth investigating what is happening here. That is what we want 
%to do now.

>>>>

\begin{abstract}
This work delves into the complex interplay between the q-boson model,
a quantum integrable system, and classical integrable systems like the
Toda and KP hierarchies. By analyzing scalar products of Bethe states
and correlation functions, we investigate their connections to tau
functions and explore Schur polynomial expansions. This exploration
unravels the intricate relationship between quantum and classical
integrable systems.
\end{abstract}

@@

Additionally, we explore other correlation functions within this model
and examine how they fit into the classical integrable context. It is
noteworthy that even when the object is not a tau function itself, it
may have an interesting expansion in terms of different classes of
Schur polynomials, highlighting some of its combinatorial
properties\footnote{Numerous results involving Schur and
Hall-Littlewood polynomials have been confirmed using the
software~\cite{Araujo:2024piv}.}. We also investigate some of these
expansions in this paper.

@@

Furthermore, according to Zabrodin~\cite{Zabrodin:2010ii}, see
citations therein, we also know that under the rescaling \(t_k \to
T_k/ \hbar\), \(t'_k \to T_{-k}/ \hbar\) and \(N = T_0/ \hbar\), we
obtain the dispersion limit tau function
\begin{equation}
  F_0(\bm{T}) = \log \mathcal{I}_{T_0}(\bm{T}, \bm{T}') + \mathcal{O}(\hbar)\; ,
\end{equation}
that is a free energy, from the viewpoit of the matrix integral
partition function.

@@

In this work, we have explored the connections between a quantum
integrable system, the q-boson model, and solutions of a classical
integrable system, the Toda hierarchy. Our investigation has extended
some early findings in this field and has also unveiled new research
avenues, which we aim to explore in future studies.  Let us now delve
into some of these promising directions.

In Section 3, we delved into various facets of the phase model. We
presented the scalar product of two off-shell Bethe states as an
elementary example of the Toda system tau function. Additionally, we
revealed an intriguing alternative portrayal of these scalar products
using a matrix integral framework, which corresponds to an ensemble of
Coulomb particles. This discovery opens up promising avenues for
further inquiry. Specifically, it would be interesting to delve deeper
into this subject and explore whether the matrix model description
offers valuable insights into the phase model and its Bethe
roots. Such an investigation promises to illuminate the underlying
dynamics of the phase model and its relationship with classical
integrable systems.

We have elucidated how various correlation functions align with this
framework, revealing a remarkably rich structure. Notably, we explored
a mapping between the phase model data and the vertex operator
representation of free bosons. An intriguing avenue for future
research is to delve into the properties of the hypergeometric tau
functions uncovered in Section 3.3.1. Investigating these functions
promises to yield valuable insights, and comparing them with existing
literature will further enhance our understanding of their
characteristics and implications within the context of integrable
systems.

In Section 4, we addressed the same problem within the context of the
q-boson model. Initially, we derived a determinant formula for the
q-boson, followed by a discussion on its expression as a quotient of
two phase model tau functions. This result establishes a connection
between the q-boson and phase model scalar product. Furthermore, we
explored three different expansions for these scalar products. Notably, we
observed that they can be expanded in terms of Big Schur and
supersymmetric tau functions. Consequently, we introduced a new set of
coordinates and demonstrated that it can be expressed as a Toda tau
function in terms of this new coordinate set.

One of our most pressing challenges lies in elucidating the intricate
connection between our findings and the Ablowitz-Ladik
hierarchy. Since this hierarchy arises as a reduction of the Toda
hierarchy, it becomes imperative to understand how we can capture the
structure of the former using the results we've derived. Given that
the q-boson model can be seen as a quantization of the Ablowitz-Ladik
equation, it stands to reason that we should observe some resemblance of
this classical problem within the framework of the q-boson
system. Understanding this relationship promises valuable insights
into the interplay between classical and quantum integrable systems.

We hope to address some of these challenges in future publications.

\subsection*{Acknowledgments}
This work is supported by Fapesp through grant \textbf{2022/06599-0}.
